---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Tornadoes over chase country

1. I suggest that smoother terrain is more conducive for tornado genesis especially when the nocturnal low-level jet is most pronounced due to local enhancement of helicity. The effect is most pronounced later in the day when the low-level jet arrives. The effect is most pronounced during months when the large-scale dynamics is weaker (May and June).

a. Use only data after the year in which population density is no longer a significant factor in reports.
b. Use only May & June tornadoes
c. Use only EF1+ or EF2+ tornadoes
d. The tornado data are nested: they come in outbreaks. This needs to be considered (e.g., only choose 1 per day).

2. The soil moisture gradient (dry-west and wet-east) portends seasonal tornado activity. Compute KBDI at two stations (Garden City-west and Salina-east) and then take the difference. Correlate the difference with tornado activity over chase country at the monthly/seasonal timescale.

1. The influence of terrain on tornado genesis across chase country

Descriptive study

Tornado start locations
```{r}
Torn.sf <- sf::st_read(dsn = here::here("data", "1950-2020-torn-initpoint"), 
                       layer = "1950-2020-torn-initpoint") |>
  dplyr::filter(mag >= 1, mo %in% 1:12) |>
  dplyr::mutate(Date = as.Date(date),
                DateTime = as.POSIXct(paste(yr, mo, dy, time), 
                                      format = "%Y%m%d%H:%M:%S"),
                Hour = lubridate::hour(DateTime),
                Year = lubridate::year(DateTime)) |>
  dplyr::group_by(Date) |>
  dplyr::sample_n(1)

#f <- here::here("data", "1950-2020-torn-initpoint")
#Torn.v <- terra::vect(f)

Torn.v <- terra::vect(Torn.sf)
class(Torn.v)
dim(Torn.v)
```

Download tiles from http://viewfinderpanoramas.org/dem3.html#nam

```{r, eval=FALSE}
dem_url <- "http://viewfinderpanoramas.org/dem3/J14.zip"
download.file(dem_url,
              destfile = here::here("data", "J14.zip"))
unzip(zipfile = here::here("data", "J14.zip"),
      exdir = here::here("data"))
```

```{r}
library(terra)
Elev1.r <- terra::rast(here::here("data", "J14", "N36W099.hgt"))
# plot(Elev1.r) # check
# dim(Elev1)
Elev2.r <- terra::rast(here::here("data", "J14", "N37W099.hgt"))
Elev3.r <- terra::rast(here::here("data", "J14", "N38W099.hgt"))
Elev4.r <- terra::rast(here::here("data", "J14", "N36W100.hgt"))
Elev5.r <- terra::rast(here::here("data", "J14", "N37W100.hgt"))
Elev6.r <- terra::rast(here::here("data", "J14", "N38W100.hgt"))
Elev7.r <- terra::rast(here::here("data", "J14", "N36W101.hgt"))
Elev8.r <- terra::rast(here::here("data", "J14", "N37W101.hgt"))
Elev9.r <- terra::rast(here::here("data", "J14", "N38W101.hgt"))


Elev.m <- terra::mosaic(Elev1.r, Elev2.r, Elev3.r, 
                        Elev4.r, Elev5.r, Elev6.r,
                        Elev7.r, Elev8.r, Elev9.r)
```

```{r}
library(tmap)
tmap_mode("view")
tm_shape(Elev.m) +
  tm_raster(palette = terrain.colors(24), alpha = .5) +
  tm_graticules(lines = FALSE)
```

Terrain roughness
```{r}
TR.m <- terra::terrain(Elev.m, 
                       v = 'roughness',
                       neighbors = 4)
Slope.m <- terra::terrain(Elev.m, 
                       v = 'slope',
                       neighbors = 4)
```

Population grid https://sedac.ciesin.columbia.edu/data/set/gpw-v4-population-density-rev11/
```{r, eval=FALSE}
file <- "gpw-v4-population-density-rev11_2020_30_sec_tif.zip"
unzip(zipfile = here::here("data", file),
      exdir = here::here("data"))

Pop.r <- terra::rast(here::here("data", 
                                "gpw_v4_population_density_rev11_2020_30_sec.tif"))
Pop.r <- terra::crop(Pop.r, Elev.m)
```

```{r}
tm_shape(log2(Slope.m)) +
  tm_raster(palette = "Blues", alpha = 1) +
  tm_graticules(lines = FALSE) +
tm_shape(Torn.sf) +
  tm_dots() +
  tm_layout(legend.outside = TRUE)
```

Crop the tornadoes to the elevation mosaic then extract the elevation and population and terrain roughness at the tornado locations. Missing values will occur for roughness and population for locations on the borders of the domain.
```{r}
Torn.v <- terra::crop(Torn.v, Elev.m)

Elev.df <- terra::extract(Elev.m, Torn.v)
mean(Elev.df[, 2], na.rm = TRUE)

TR.df <- terra::extract(TR.m, Torn.v)
mean(TR.df[, 2], na.rm = TRUE)

Pop.df <- terra::extract(Pop.r, Torn.v)
mean(Pop.df[, 2], na.rm = TRUE)
```

Random sample of tornado locations over the domain. Background locations. https://rspatial.org/terra/sdm/index.html
```{r}
Domain.e <- terra::ext(Elev.m)
class(Domain.e)
```

Background data characterize the range of environmental conditions in the region. In this sense, background is the same, irrespective of where the tornadoes have originated. Background data establishes the environmental domain of the study, whilst the tornado locations establish under which conditions a tornado is more likely to be present than on average. 

```{r}
Avg = NULL
for(i in 1:1000){
Background.v <- terra::spatSample(Domain.e, 
                                  size = nrow(Torn.v),
                                  method = "random",
                                  lonlat = TRUE,
                                  as.points = TRUE)
Avg[i] <- mean(terra::extract(TR.m, Background.v)[, 2], na.rm = TRUE)
}
hist(Avg)
```

Create a single data frame with tornado locations (labeled 1) and background locations (labeled 0) that includes elevation, population, and terrain roughness.
```{r}
Torn.df <- data.frame(Location = 1, 
                      Elevation = Elev.df[, 2], 
                      Roughness = TR.df[, 2], 
                      Population = Pop.df[, 2])

Background.v <- terra::spatSample(Domain.e, 
                                  size = nrow(Torn.v),
                                  method = "random",
                                  lonlat = TRUE,
                                  as.points = TRUE)
ElevB.df <- terra::extract(Elev.m, Background.v)
TRB.df <- terra::extract(TR.m, Background.v)
PopB.df <- terra::extract(Pop.r, Background.v)
Background.df <- data.frame(Location = 0, 
                            Elevation = ElevB.df[, 2], 
                            Roughness = TRB.df[, 2], 
                            Population = PopB.df[, 2])

TornBack.df <- rbind(Torn.df, Background.df)
```

Logistic regression
```{r}
model1 <- glm(Location ~ Population + Elevation + Roughness,
              family = "binomial", data = TornBack.df)
summary(model1)

model2 <- glm(Location ~ Elevation + Roughness,
              family = "binomial", data = TornBack.df)
summary(model2)

p <- predict(model2, newdata = Torn.df, type = "response")
b <- predict(model2, newdata = Background.df, type = "response")

e <- predicts::pa_evaluate(p = p, a = b)
e@stats
```

```{r}
remotes::install_github("rspatial/predicts")

predictors <- c(Elev.m, TR.m)
names(predictors) <- c("Elevation", "Roughness")

pg <- predict(predictors, model2, type = "response")
plot(pg)
```

Create supercells https://github.com/Nowosad/supercells
```{r}
pg1 <- supercells::supercells(pg, k = 200, compactness = .1)

tm_shape(pg1) +
  tm_fill(col = "lyr1", title = "Probability of tornado genesis") +
  tm_layout(legend.outside = TRUE) +
  tm_graticules(lines = FALSE) +
tm_shape(Torn.sf) +
  tm_dots() 
```

Create start and end spatial vectors
```{r}
TornStart.v <- terra::vect(as.data.frame(Torn.v), 
                           geom = c("slon", "slat"),
                           crs = crs(Torn.v)) |>
  terra::crop(Elev.m)

TornEnd.v <- terra::vect(as.data.frame(Torn.v), 
                         geom = c("elon", "elat"),
                         crs = crs(Torn.v)) |>
  terra::crop(Elev.m)

mean(terra::extract(TR.m, TornStart.v)[, 2], na.rm = TRUE)
mean(terra::extract(TR.m, TornEnd.v)[, 2], na.rm = TRUE)
```

No difference is roughness for start and end locations.


Daily weather data
```{r}
Salina.df <- readr::read_csv(file = here::here("data", "Salina_Kansas.csv")) |>
  dplyr::mutate(Year = lubridate::year(DATE)) |>
  dplyr::filter(Year >= 2000) |>
  dplyr::select(DATE, Year, PRCP, TMAX, TMIN, TAVG)

GardenCity.df <- readr::read_csv(file = here::here("data", "GardenCity_Kansas.csv"))


```
https://www.ncei.noaa.gov/maps/daily-summaries/